---
title: "Flow of Funds in the United States"
author: "Manolis Chatzikonstantinou"
date: '2022-07-23'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Data from the Federal Reserve

Load the packages and connect to the database API

```{r echo=FALSE}
pacman::p_load(fredr,extrafont,purrr,tidyverse,ggplot2,gt,glue)



fredr::fredr_set_key("3bc35d4a8b94bd0f461a3a61f3d804ba")
```

```{r}

data<-fredr_source_releases(
  source_id = 1L
)


# Create a gt table based on preprocessed
# `sp500` table data
data %>% head(5) %>% select(name,link) %>% set_names("Table","Link") %>% 
  gt()  %>%
  tab_header(
    title = "Data available in the FRED",
    subtitle = "Provided by the Board of Governors of the Fed"
  ) %>%
  tab_options(
    container.width = pct(85),
    container.overflow.x = T,
    table.align = "center",
    table.width = pct(85),
    table.layout = "auto"
  )
```

We care about the following table:
```{r}


# Create a gt table based on preprocessed
# `sp500` table data
data %>% slice(10) %>% select(name,link) %>% set_names("Table","Link") %>% 
  gt()  %>%
  tab_header(
    title = "Data available in the FRED",
    subtitle = "Provided by the Board of Governors of the Fed"
  ) %>%
  tab_options(
    container.width = pct(85),
    container.overflow.x = T,
    table.align = "center",
    table.width = pct(85),
    table.layout = "auto")
```


You may be interested about other data published in the database. If you would like to learn more, you can make use of the [Categories endpoint](https://fred.stlouisfed.org/docs/api/fred/#Categories) of the FRED API. In particular, FRED series are assigned categories. Each FRED category is assigned an integer identifier. For example:
  - "Population, Employment, & Labor Markets" (category_id = 10)
  - "National Accounts" (category_id = 32992)
  - "Production & Business Activity" (category_id = 1)

Then, you can see what are the subcategories related to each category:

```{r}
fredr_category_children(category_id = 1L)
```

The function `fredr_category_series()` returns detailed information for the FRED series belonging to the category specified by `category_id`. For example, to get the top 10 annual series in the "Population, Employment, & Labor Markets" category by popularity:

```{r}
fredr_category_series(
  category_id = 1L, limit = 10L, order_by = "popularity",sort_order="desc")
```

You can find also the more recent releases

```{r}
fredr_releases_dates()

```

The function fredr_release_series() returns a set of series belonging to the FRED release specified by release_id. The data returned is a tibble where each row represents a series in the release specified. For example, to get series in the Employment Cost Index release:



```{r}
fredr_release_series(release_id = 11L)
```

Note the parameters available to filter series belonging to a release:

```{r}
fredr_release_series(
  release_id = 10L,
  filter_variable = "frequency",
  filter_value = "Monthly",
  order_by = "popularity",
  sort_order = "desc",
  limit = 10L
)
```
Get the categories for a FRED series
The fredr_series_categories() function returns a list of categories for the series specified by series_id. The data returned is a tibble in which each row represents a category that the series belongs to. For example, to get the categories for the UNRATE series:

fredr_series_categories(series_id = "UNRATE")
