---
title: "T tests in Class"
author: "Manolis Chatzikonstantinou"
date: "12/1/2021"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Question

A leakage test was conducted to determine the effectiveness of a seal designed to keep the inside of a plug airtight.
An air needle was inserted into the plug, and the plug and needle were placed under water.
The pressure was then increased until leakage was observed.
Let $X$ equal the pressure in pounds per square inch.
Assume, that the distribution of $X$ is $N\left(\mu, \sigma^{2}\right)$.
The following $n=10$ observations of $X$ were obtained: $$
\begin{array}{llllllllll}3.1 & -3.3 & -4.5 & -2.8 & 3.5 & 3.5 & -3.7 & 4.2 & -3.9 & 3.3\end{array}
$$.

Use the observations to

1.  Find a point estimate of $\mu$.

2.  Find a point estimate of $\sigma$.

3.  Find a $95 \%$ confidence interval for $\mu$ .

4.  Can you reject that the mean is different than zero, under the restriction that your test has a probability of Type 1 error equal to 0.05.

## T - tests in R

```{r}
data_final<-c(3.1,-3.3,-4.5,-2.8,3.5,3.5,-3.7,4.2,-3.9,3.3)
```

For item 1: I need to calculate the sample mean.

```{r}
mean(data_final)
```

For item 2: I need to calculate the sample standard deviation, since the sample standard deviation is an unbiased estimator of $\sigma$.

```{r}
sd(data_final)
```

For item 3, I need the value of the t-score for 9 degrees of freedom and $\alpha/2$ equal to 0.025.

```{r}
tscore = -qt(0.025,df = 9)
tscore
```

So a lower bound is

```{r}
mean(data_final) - tscore*sd(data_final)/sqrt(10)
```

For an upper bound

```{r}
mean(data_final) + tscore*sd(data_final)/sqrt(10)
```

For item 4: I need to do a t test

```{r}
t.test(data_final)
```

I do not reject the initial hypothesis becuase the p-value is bigger than 5 percent.
