---
title: "Replicating Money Banking Figures and Tables"
author: "Manolis Chatzikonstantinou"
date: "4/17/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Fredr recap

Load the packages and connect to the database API

```{r}
pacman::p_load(fredr,extrafont,purrr,tidyverse,ggplot2)
fredr::fredr_set_key("3bc35d4a8b94bd0f461a3a61f3d804ba")
```



# Find a series

Now we would like to find the series related to commercial paper;

```{r}
popular_funds_series <- fredr_series_search_text(
    search_text = "commercial paper",
    order_by = "popularity",
    sort_order = "desc"
)

popular_funds_series %>% filter(substr(id,1,10) == "BOGZ1FL893") 

```


Now we would like to find the series related to commercial paper;


```{r}
# All Sectors; Corporate Equities; Asset, Market Value Levels
popular_funds_series <- fredr_series_search_text(
    search_text = "equity",
    order_by = "popularity",
    sort_order = "desc"
)

popular_funds_series %>% filter(substr(id,1,9) == "BOGZ1LM89")  %>% select(id,title)

```




```{r}
# All Sectors; Corporate Equities; Asset, Market Value Levels
popular_funds_series <- fredr_series_search_text(
    search_text = "mortgage",
    order_by = "popularity",
    sort_order = "desc"
)

popular_funds_series %>% filter(substr(id,1,9) == "BOGZ1FL89")  %>% select(id,title)

```


```{r}
# All Sectors; Corporate Equities; Asset, Market Value Levels
popular_funds_series <- fredr_series_search_text(
    search_text = "bonds",
    order_by = "popularity",
    sort_order = "desc"
)

popular_funds_series %>% filter(substr(id,1,9) == "BOGZ1FL89")  %>% select(id,title)

```




```{r echo=FALSE}
# All Sectors; Corporate Equities; Asset, Market Value Levels
popular_funds_series <- fredr_series_search_text(
    search_text = "securities",
    order_by = "popularity",
    sort_order = "desc"
)

popular_funds_series %>% filter(substr(id,1,9) == "BOGZ1LM89")  %>% select(id,title)
# 59 BOGZ1FL894122005A All Sectors; Total Debt Securities; Liability, Level     
# 28 BOGZ1FL894022005Q All Sectors; Total Debt Securities; Asset, Level                                             
#15 BOGZ1FL894104005A All Sectors; Debt Securities and Loans; Liability, Level 
# 75 BOGZ1FA893061105Q All Sectors; Treasury Securities; Asset, Transactions                                        
# 24 BOGZ1FL893061105Q All Sectors; Treasury Securities; Asset, Level  



#BOGZ1LM893064105A	All Sectors; Corporate Equities; Asset, Market Value Levels
#BOGZ1FL893163005A	All Sectors; Corporate and Foreign Bonds; Liability, Level		
#BOGZ1FL893063005A	All Sectors; Corporate and Foreign Bonds; Asset, Level

# BOGZ1FL893065125Q All Sectors; Home Equity Loans; Asset, Level                                                        

     
     

```



```{r}
# All Sectors; Corporate Equities; Asset, Market Value Levels
popular_funds_series <- fredr_series_search_text(
    search_text = "government",
    order_by = "popularity",
    sort_order = "desc"
)

popular_funds_series %>% filter(substr(id,1,9) == "BOGZ1LM89")  %>% select(id,title)

```

## Chapter 2

Table 1 Principal Money Market Instruments
```{r data}

pacman::p_load(fredr,extrafont,purrr,tidyverse,ggplot2)
fredr::fredr_set_key("3bc35d4a8b94bd0f461a3a61f3d804ba")

```


# Get the data

We download the data and then name each series in the way, we present them in the table:
```{r}
# Corporate stocks (market value)-BOGZ1LM893064105A
# Residential mortgages-BOGZ1FL893065105A
# Corporate bonds-BOGZ1FL893163005A
# U.S. government securities (marketable long-term)-BOGZ1FL313161205A
# U.S. government agency securities - BOGZ1FL893161705A
#State and local government bonds - SLGMSOA027N
# Bank loans- BOGZ1FL793068005A
# Consumer loans-HNOCCIA027N
# Commercial and farm mortgages-BOGZ1FL893065603A+BOGZ1FL893065505A


datatable2<-bind_rows(fredr(series_id = "BOGZ1LM893064105A",observation_start = as.Date("1990-01-01")) %>% 
  mutate(sname = "Corporate stocks (market value)") %>% mutate(value  = value/1000) %>% select(1,3,sname),
  fredr(series_id = "BOGZ1FL893065105A",observation_start = as.Date("1990-01-01")) %>% 
  mutate(sname = "Residential mortgages") %>% mutate(value  = value/1000) %>% select(1,3,sname),
  fredr(series_id = "BOGZ1FL893163005A",observation_start = as.Date("1990-01-01")) %>% 
  mutate(sname = "Corporate bonds") %>% mutate(value  = value/1000) %>% select(1,3,sname),
  fredr(series_id = "BOGZ1FL313161205A",observation_start = as.Date("1990-01-01")) %>% 
  mutate(sname = "U.S. government securities (marketable long-term)") %>% mutate(value  = value/1000) %>% select(1,3,sname),
  fredr(series_id = "BOGZ1FL893161705A",observation_start = as.Date("1990-01-01")) %>% 
  mutate(sname = "U.S. government agency securities") %>% mutate(value  = value/1000) %>% select(1,3,sname),
  fredr(series_id = "SLGMSOA027N",observation_start = as.Date("1990-01-01")) %>% 
  mutate(sname = "State and local government bonds") %>% mutate(value  = value/1000) %>% select(1,3,sname),
    fredr(series_id = "BOGZ1FL793068005A",observation_start = as.Date("1990-01-01")) %>% 
  mutate(sname = "Bank loans") %>% mutate(value  = value/1000) %>% select(1,3,sname),
    fredr(series_id = "HNOCCIA027N",observation_start = as.Date("1990-01-01")) %>% 
  mutate(sname = "Consumer loans") %>% mutate(value  = value/1000) %>% select(1,3,sname),
   inner_join(fredr(series_id = "BOGZ1FL893065603A",observation_start = as.Date("1990-01-01")),fredr(series_id = "BOGZ1FL893065505A",observation_start = as.Date("1990-01-01")),by="date") %>% mutate(value = (value.x+ value.y)/1000,sname="Commercial and farm mortgages") %>% select(date,value,sname))


```



## Including Tables

You can also embed tables, for example:

```{r pressure, echo=FALSE, eval=FALSE}
# install.packages("gt")
library(gt)
library(tidyverse)

# Define the start and end dates for the data range
dates <- c("1990-01-01","2000-01-01","2010-01-01","2016-01-01")
dates<- as.Date(dates)
# Create a gt table based on preprocessed
#
my_table<-datatable2 %>% filter(date %in% dates)%>% mutate(year=substr(date,1,4)) %>% 
  select(2,3,4)  %>% 
   mutate( year = paste("Amount ($ billions end of year)",year,sep = " ")) %>%
  pivot_wider(names_from = year,values_from = value) %>%
  gt() %>% tab_header(title = "Table 2: Principal Capital Market Instruments")  %>% 
  fmt_number( # Another column (also numeric data)
    columns = 2:5, # What column variable? BOD$demand
    decimals = 1 # I want this column to have one decimal place
  )%>%cols_label(sname="Type of instrument") %>% 
  fmt_markdown(columns = everything())  %>%
  cols_width(sname ~ px(200),
             everything() ~ px(120))


gtsave(my_table,filename = "week1graphs/table2.jpg")
```
 
```{r include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
knitr::include_graphics("my_table.pdf")
```


